---
title: "The effects of severe weather events on population health and economics between 1996 and 2011."
author: "Stijn Servaes"
date: "06 July 2017"
output: 
  html_document:
    keep_md: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(plyr)
library(gridExtra)
library(reshape2)
options(scipen=999)
```

## 1. SYNOPSIS

The National Oceanic and Atmospheric Administration (NOAA) maintains a public database for storm events. The data that is collected here displays information on the type of storm and details like location, date, estimates for damage to property as well as the number of human victims of the storm. In this report we investigate which type of events are the most harmful to the population and economically between the period of 1996 and 2011.  

From this analyis it is clear that tornadoes have the highest number of victims, causing the highest number of injuries and the second highest number of fatalities, closely following excessive heat events. Besides tornadoes, other major weather types having a high number of victims were excessive heat events and floods.  
In terms of economic costs however, weather types that occur on a larger scale and have a longer timespan than when compared to tornadoes, seem to have more impact. The leading weather type in terms of costs were floods, having the highest cost in property damage and coming in second in terms of crop damage, being preceded by droughts.  
Furthermore, it appears that the costs for crop damage do not seem to be a major factor in the total costs of damages caused by weather events, as these are majorily governed by costs for property damage. Besides floods, other major weather types having a high economical impact were hurricanes/typhoons and storm surges.

***

## 2. INTRODUCTION

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

***

## 3. DATA

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined:

* [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

***

## 4. DATA PROCESSING

Download, unzip and load data into a dataframe called `storm`.
As data in the earlier years of the dataset is likely to be skewed due to a lack of good records, a subset of the more recent years was made. This date was arbitrarily set at 1996-01-01.

```{r storm, cache = TRUE}
# Load in the data
setwd("/Users/sservaes/datasciencecoursera/RepData_PeerAssessment2")
if(!file.exists("StormData.csv.bz2")) {
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "StormData.csv.bz2")
}
storm <- read.csv("StormData.csv.bz2")

# Subset the dataset by removing the data before 1996-01-01
storm$BGN_DATE <- as.Date(storm$BGN_DATE, format = "%m/%d/%Y", "%Y")
storm_select <- storm[which(storm$BGN_DATE >= "1996-01-01"),]
```

In order to identify which severe weather type had the most influence on population health, the data was further processed as follows:

* A new dataframe `EVTYPE_pop` was made to summarise variables according to `EVTYPE`.
* The sum of `FATALITIES` was made according to `EVTYPE` in a variable `fatalities`.
* The sum of `INJURIES` was made according to `EVTYPE` in a variable `injuries`.
* `FATALITIES` and `INJURIES` were summed together according to `EVTYPE` in a variable `affected_population`.
* In order to identify the worst severe weather types in respect to population health, a variable `EVTYPE_pop_worst` was made which grouped all the weather types that affected more than 1000 people (injuries or fatalities).

``` {r population, cache = TRUE}
# Summarise the dataset by fatalities and injuries.
EVTYPE_pop <- ddply(storm_select, ~EVTYPE, summarise, 
                    fatalities = sum(FATALITIES), 
                    injuries = sum(INJURIES), 
                    affected_population = sum(FATALITIES)+sum(INJURIES))

# Determine the worst weather types (those that affect more that 1000 people).
EVTYPE_pop_worst <- EVTYPE_pop[which(EVTYPE_pop$affected_population>1000),]


# Melt the data according to EVTYPE by fatalities and injuries.
EVTYPE_pop_melt <- melt(EVTYPE_pop_worst, id.vars = "EVTYPE", 
                        measure.vars =  c("fatalities", "injuries"), 
                        variable.name="severity", value.name="health")
```

In order to identify which severe weather type had the most influence on economics, the data was further processed as follows:

* Total damages were calculated taking the exponential into account.
* In order to identify the worst severe weather types in respect to economics, a variable `EVTYPE_eco_worst` was made which grouped all the weather types that had a total cost higher than 5 billion USD (property or crops damage).

``` {r economic, cache = TRUE}
# Calculated total damage to property.
storm_select$PROPDMGTOT <- storm_select$PROPDMG
storm_select$PROPDMGTOT[which(storm_select$PROPDMGEXP=="K")] <- storm_select$PROPDMG[which(storm_select$PROPDMGEXP=="K")] * 1000
storm_select$PROPDMGTOT[which(storm_select$PROPDMGEXP=="M")] <- storm_select$PROPDMG[which(storm_select$PROPDMGEXP=="M")] * 1000000
storm_select$PROPDMGTOT[which(storm_select$PROPDMGEXP=="B")] <- storm_select$PROPDMG[which(storm_select$PROPDMGEXP=="B")] * 1000000000

# Calculated total damage to crops.
storm_select$CROPDMGTOT <- storm_select$CROPDMG
storm_select$CROPDMGTOT[which(storm_select$CROPDMGEXP=="K")] <- storm_select$CROPDMG[which(storm_select$CROPDMGEXP=="K")] * 1000
storm_select$CROPDMGTOT[which(storm_select$CROPDMGEXP=="M")] <- storm_select$CROPDMG[which(storm_select$CROPDMGEXP=="M")] * 1000000
storm_select$CROPDMGTOT[which(storm_select$CROPDMGEXP=="B")] <- storm_select$CROPDMG[which(storm_select$CROPDMGEXP=="B")] * 1000000000

# Summarise the dataset by property and crop damage.
EVTYPE_eco <- ddply(storm_select, ~EVTYPE, summarise, property_damage = sum(PROPDMGTOT), crop_damage = sum(CROPDMGTOT), total_costs = sum(PROPDMGTOT)+sum(CROPDMGTOT))

# Determine the worst weather types for economics (those that have a total cost higher than 5 billion USD).
EVTYPE_eco_worst <- EVTYPE_eco[which(EVTYPE_eco$total_costs>5000000000),]

# Melt the data according to EVTYPE by property_damage and crop_damage
EVTYPE_eco_melt <- melt(EVTYPE_eco_worst, id.vars = "EVTYPE", 
                        measure.vars =  c("crop_damage", "property_damage"), 
                        variable.name="damage_type", value.name="damage_cost")
```

***

## 5. RESULTS

### 5.1 Population Health

In order to identify which types of events (as indicated in the ð™´ðš…ðšƒðšˆð™¿ð™´ variable) are most harmful with respect to population health in the United States between 1996 and 2011, the variables `fatalities` and `injuries` were plotted between that time period.

``` {r plot_population}
# Data used for the plot.
print(EVTYPE_pop_worst, row.names = FALSE)

# Generate the combined plot.
plot_pop <- ggplot(EVTYPE_pop_melt, aes(x = reorder(EVTYPE, -health), y = health/1000)) +
        geom_bar(stat = "identity", aes(fill = severity)) +
        theme(axis.text.x = element_text(angle = 45, size=8, hjust = 1, vjust = 1), 
              axis.title.x = element_blank()) + 
        ggtitle("Fatalities and Injuries (1996 - 2011)") + 
        ylab("Number of people (in thousands)") +
        scale_fill_manual(values=c("cornflowerblue", "coral1"), 
                          labels = c("Fatalities", "Injuries"), 
                          name = "Severity")
plot_pop
```

From the plots it is clear that in terms of fatalities the following 3 weather types have the highest numbers. 
```{r}
table_fatalities <- EVTYPE_pop[order(-EVTYPE_pop$fatalities),][1:3,1:2]
print(table_fatalities, row.names = FALSE)
```

The highest amount of injuries were seen in the following 3 weather types:

```{r}
table_injuries <- EVTYPE_pop[order(-EVTYPE_pop$injuries),][1:3,c(1,3)]
print(table_injuries, row.names = FALSE)
```

The highest amount of injuries and fatalities combined were seen in the following 3 weather types:

```{r}
table_combined <- EVTYPE_pop[order(-EVTYPE_pop$injuries),][1:3,c(1,4)]
print(table_combined, row.names = FALSE)
```

From this analysis it is clear that between 1996 and 2011, weather type `r table_combined$EVTYPE[1]` affected the highest amount of people, with in total `r table_combined$affected_population[1]` people affected. In terms of injuries weather type `r table_injuries$EVTYPE[1]` had the highest amount of people with `r table_injuries$injuries[1]` injuries. In terms of fatalities weather type `r table_fatalities$EVTYPE[1]` had the highest amount of people with `r table_fatalities$fatalities[1]` fatalities.

### 5.2 Economic consequences

In order to identify which types of events (as indicated in the ð™´ðš…ðšƒðšˆð™¿ð™´ variable) are most harmful with respect to population health in the United States between 1996 and 2011, the variables `fatalities` and `injuries` were plotted between that time period.

``` {r plot_economics}
# Data used for the plot.
EVTYPE_eco_worst <- EVTYPE_eco_worst[order(-EVTYPE_eco_worst$total_costs),]
print(EVTYPE_eco_worst, row.names = FALSE)

# Generate the combined plot.
plot_eco <- ggplot(EVTYPE_eco_melt, aes(x = reorder(EVTYPE, -damage_cost), y = damage_cost/1000000000)) +
        geom_bar(stat = "identity", aes(fill = damage_type)) +
        theme(axis.text.x = element_text(angle = 45, size=8, hjust = 1, vjust = 1), axis.title.x = element_blank()) + 
        ggtitle("Economic Costs (1996 - 2011)") + 
        ylab("USD (in billions)") +
        scale_fill_manual(values=c("cornflowerblue", "coral1"), 
                          labels = c("Damage to crops", "Damage to property"), 
                          name = "Damage type")
plot_eco
```

From the plots it is clear that in terms of property damage the following 3 weather types have the highest numbers. 
```{r}
table_property <- EVTYPE_eco[order(-EVTYPE_eco$property_damage),][1:3,1:2]
print(table_property, row.names = FALSE)
```

The highest amount of crop damage were seen in the following 3 weather types:

```{r}
table_crops <- EVTYPE_eco[order(-EVTYPE_eco$crop_damage),][1:3,c(1,3)]
print(table_crops, row.names = FALSE)
```

The highest amount of total costs were seen in the following 3 weather types:

```{r}
table_eco <- EVTYPE_eco[order(-EVTYPE_eco$total_costs),][1:3,c(1,4)]
print(table_eco, row.names = FALSE)
```


From this analysis it is clear that between 1996 and 2011, weather type `r table_eco$EVTYPE[1]` had the highest amount of costs, with in total `r round(table_eco$total_costs[1]/1000000000, 2)` billion USD. In terms of property damage weather type `r table_property$EVTYPE[1]` had the highest cost with `r round(table_property$property_damage[1]/1000000000, 2)` billion USD. In terms of crop damage weather type `r table_crops$EVTYPE[1]` had the highest cost with `r round(table_crops$crop_damage[1]/1000000000, 2)` billion USD.